---
title: "4 Data wrangling - Report Exerise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

## Analyzing changes in soil organic matter during elevated CO2 experiments
 
In this report exercise, I will practice Open Science principles by converting a badly published data set into a data set that is usable and consequently allows for a reproducible workflow. Afterward, I will investigate the effect of increased atmospheric CO2 concentrations on soil organic matter using the previously cleaned data set.

<br>

The data is taken from *Groenigen, Kees Jan van, Xuan Qi, Craig W. Osenberg, Yiqi Luo, and Bruce A. Hungate. “Faster Decomposition Under Increased Atmospheric CO2 Limits Soil Carbon Storage.” Science 344, no. 6183 (May 2, 2014): 508–9. [https://doi.org/10.1126/science.1249534.](10.1126/science.1249534).* It is saved in the /dataraw folder. Before using it here I cleaned it manually by:

* deleting empty rows,
* deleting the first 3 rows that were empty,
* replacing spaces with "_" in column names, 
* filling the column *Experiments*,
* converting the column depth to a machine readable numeric value and
* converting *Sample dates* to a homogeneous date format.

<br>

```{r preamble, message=FALSE, warning=FALSE}
# libraries
library(dplyr); library(lubridate); library(tidyr); library(readr);library(kableExtra);library(ggplot2)
# clean data
data <- read_csv("../data/groenigen_DBS1.csv", locale = locale(encoding = "UTF-8"))[,1:11]

# convert date column to date class
data <- data |>
  mutate(Sample_date = dmy(Sample_date))
```

<br>

```{r data table, echo=FALSE}
knitr::kable(data[1:5,1:9], 
             col.names = c("Experiment","Citation","Depth","Date","Time","AmbientCO2","IncreasedCO2", "nAmbient", "nIncreased"),
             caption = "Table 1: Data overview without the columns \"Desciptions\" and \"Value treatment\".", 
             align = c("l","l","c","l",rep("c",5)))


```

### 1. Introduction
In the era of climate change, it is vital to understand the effects of increased atmospheric CO2 concentrations on ecosystems. Soil stores carbon in the form of organic and inorganic matter. Organic soil matter is cut-up dead biomass primarily contributed by plants. The growth of plants is regulated by their photosynthesis rate which can be limited or enhanced by the abundance of atmospheric CO2. 

In this report I will investigate whether increased atmospheric CO2 leads to and increasing storage of carbon in the soil. I hypothesize that (i) higher CO2 concentrations lead to a short term increase in soil carbon storage and (ii) that soil carbon will eventually reach an equilibrium at those higher concentrations and not rise further.  

<br>

### 2. Methods
We analyze the effect of CO2 on soil carbon concentrations by comparing two soil samples, one from an environment with ambient CO2 concentrations, the other with increased CO2 concentrations.

We work with data provided by *Groenigen et al. (2014)* who summarized existing literature on the responses of elevated CO2 concentrations on soil organic matter. We calculate the log-response ratio comparing ambient CO2 and elevated CO2 soil organic matter across all experiment. However, we will distinguish between different time phases after the start of the experiment. These could indicate a timely development of the response to the elevated CO2 concentrations. Also these phases could show the formation of a possible equilibrium. The phases are:

* early: start of the experiment and up to 3 years
* mid: between 3 and 6 years after the start of the experiment
* late: more than 6 years since the start of the experiment

The formula used for calculating the log-response ratio RR is with $x$ indicating the concentration of soil carbon in different scenarios:
$$RR = ln( \frac{x_{elevated}}{x_{ambient}} )$$


To get the average log-response ratio for each phase across all experiments we:

1. calculated the averages of ambient and increased CO2 conditions for every experiment with respect to the phases,
2. then calculated the log-response ratios of all existing experiment phase pairs and
3. then average over all values of each phase.
 
 
The averaging of the log-response ratios had to be adjusted, because not all phases had measured over enough time to represent all of our phases or were missing data. Consequently, we had to exclude all response ratios from the averaging process that were either not a number (NaN) and exactly 0 because one or both of the division partners were missing. A response ratio of 0 could also mean that there is exactly no effect of the increased CO2 concentrations. However, having precisely no difference is highly unlikely and could also result out of measurement errors.
<br>

```{r RR}
#  aggregate across experiments and phases
data_aggregated <- data |> 
  
  # select columns that we investigate
  select(Experiment, Time_years, mean_ambient_CO2, mean_increased_CO2) |>
  
  # group by experiment
  group_by(Experiment) |>
  
  # 1. summarize across ambient and increased CO2 values in 3 phases
  summarize(mean_ambient_CO2_early = mean(mean_ambient_CO2[Time_years < 3], na.rm = T),                        # early
            mean_ambient_CO2_mid   = mean(mean_ambient_CO2[Time_years >= 3 & Time_years <= 6], na.rm = T),     # mid
            mean_ambient_CO2_late  = mean(mean_ambient_CO2[Time_years > 6], na.rm = T),                        # late
            
            mean_increased_CO2_early = mean(mean_increased_CO2[Time_years < 3], na.rm = T),                    # early
            mean_increased_CO2_mid   = mean(mean_increased_CO2[Time_years >= 3 & Time_years <= 6], na.rm = T), # mid
            mean_increased_CO2_late  = mean(mean_increased_CO2[Time_years > 6], na.rm = T),                    # late
            
            # 2. calculate RR for every parallel observation 
            RR_early = log(mean_increased_CO2_early / mean_ambient_CO2_early),   # early
            RR_mid   = log(mean_increased_CO2_mid / mean_ambient_CO2_mid),       # mid
            RR_late  = log(mean_increased_CO2_late / mean_ambient_CO2_late)      # late
            )

# 3. average the response ratios for every phase
data_RR_means <- data_aggregated |>
  summarize(mean_RR_early = mean(RR_early[!is.nan(RR_early) & RR_early != 0], na.rm = T),        # early
            mean_RR_mid   = mean(RR_mid[!is.nan(RR_mid) & RR_early != 0], na.rm = T),            # mid
            mean_RR_late  = mean(RR_late[!is.nan(RR_late) & RR_early != 0], na.rm = T)           # late
  )
```


<br>

### 3. Results

There are clear differences between experiments and their recorded average soil carbon concentration (Table 2). From the 5 experiments shown in Table 2, there is no clear pattern visible between early, mid and late phase soil carbon concentrations. Response ratios varied from phase to phase but also experiment to experiment. Multiple experiments did not measure soil organic matter for all of our defined phases. Consequently, 32% of all response ratios are missing. 

```{r data aggregated table, echo=FALSE}
# find out the percentage of missing RR of all phases
# nan_percentage <- sum(is.nan(as.matrix(data_aggregated[,8:10]))) / prod(dim(data_aggregated[,8:10])) * 100
# -> 32%

# round values and replace NaNs with "-"
data_aggregated_table <- data_aggregated |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  mutate(across(everything(), ~ifelse(is.nan(.), "-", .)))

knitr::kable(data_aggregated_table[1:5,],  
             col.names = c("Experiment","ACO2_early","ICO2_early","ACO2_mid","ICO2_mid","ACO2_late","ICO2_late", "RR_early", "RR_mid", "RR_late"), 
             digits = 2, 
             caption = "Table 2: Cutout of the aggregated data.")

```


<br>

Through averaging over all experiments we get a more representative picture of the effect of increased atmospheric carbon on soil organic matter (Table 3). These averages show an increase in the soil organic matter due to elevated CO2 levels in the first phase (0-3 years). However already at the second phase (3-6 years) we have a negative response ratio indicating less carbon in the elevated CO2 scenario. After 6 years, the difference between elevated CO2 soil and ambient CO2 soil decreases again but stays as in phase two with slightly larger soil carbon concentrations in the ambient scenario.

```{r RR table, echo=FALSE}
knitr::kable(data_RR_means,  
             col.names = c("RR early phase","RR mid phase","RR late phase"), digits = 4,
             caption = "Table 3: Response ratios to increased CO2 concentrations over 3 phases.") 

```






<br>

### 4. Discussion
I hypothesized that elevated CO2 concentrations could lead to a short term increase in soil carbon storage, which would reach an equilibrium with higher concentrations in the elevated CO2 scenario. I did find soil carbon to increase at first due to the increased abundance of atmospheric CO2, however, after some time the soil carbon decreased again below levels of the ambient CO2 scenario. Interestingly, the soil stores less carbon with higher atmospheric carbon availability. The authors of *Groenigen et al. (2014)* argue that this could be due to an increase in the microbial activity that decomposes soil organic carbon, which I did not take into account while formulating my hypotheses. These microbes respirate soil organic carbon faster than it is produced in elevated CO2 conditions, leading to an equilibrium below the levels of the ambient scenario. The global increase in temperature could even further exacerbate this shift to a decrease of soil organic carbon storage. This could have dire consequences as soils would not act as a carbon sink that is able to keep up with rising CO2 concentrations, according to my results.

<br>
