---
title: "5 Air Quality - Report Exercise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

## Telling a story from data

target variable : Ozone, think about patterns and relationships

Must do:
- description of the airquality data set: https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality
- A research question: predictability using Wind/Solar.R/Temp or is there a weekly cycle,
- At least three statistical metrics from your dataset that aid you in answering your question (e.g., mean values, ranges, etc.).
- At least three publishable figures or tables that show important relationships that aid you in answering your question (e.g., outliers, temporal patterns, scatterplots, etc.).
- Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?
- max 400 words (aka 700)

```{r data}
# libraries
library(dplyr); library(lubridate); library(tidyr); library(readr); library(stringr); library(purrr)

# data
airquality <- datasets::airquality

plot(airquality$Ozone, airquality$Temp)
plot(airquality$Ozone, airquality$Solar.R)
plot(airquality$Ozone, airquality$Wind)

hist(airquality$Wind)


```

