---
title: "5 Air Quality - Report Exercise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,messages=FALSE, warning=FALSE)
rm(list = ls())
```

## Telling a story from data


### Notes
- At least three statistical metrics from your dataset that aid you in answering your question (e.g., mean values, ranges, etc.).
- At least three publishable figures or tables that show important relationships that aid you in answering your question (e.g., outliers, temporal patterns, scatterplots, etc.).
- Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?
- max 400 words (aka 700)
- Do I need to scale for lm()? Kinda not but could be nice to try out

### 1. Introduction
Ground level ozone is a threat to human health as it can cause lung tissue damage and negatively affect preexisting respiratory diseases. Ozone forms when pollutants from traffic or industry like nitrous oxide react with solar shortwave radiation. Therefore, ozone is highly abundant in populated urban areas like New York. 

In this report exercise, I set out to answer the following research questions regarding the timely distribution of dangerous ozone exposure in New York in 1973, as well as the predictability of ozone using other environmental variables:

* 1. Is there a seasonal and weekly pattern in the ground level ozone concentration in New York in 1973?
* 2. Can ozone concentrations be sufficiently predicted using other meteorological variables like wind speed, solar shortwave radiation and temperature?

I hypothesize that Ozone concentrations are higher in the summer due increased solar radiation and during the week due to increased traffic and industry activity. Furthermore, ozone should be best predicted by solar radiation as it is integral to its formation. 

<br>

### 2. Methods
To answer these research questions, I investigated the public dataset [*airquality*](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality) with data from May first to September 30th 1973 (Table 1).

```{r table data, echo= FALSE}
library(kableExtra)
table_data = data.frame(Variable = c("Ozone", "Solar radiation", "Wind speed", "Temperature"),
                        Unit = c("ppb", "langleys", "mph", "°C"),
                        Site = c("Roosevelt Island", "Central Park", "LaGuardia Airport", "LaGuardia Airport"),
                        Method = c("average from 13:00 to 15:00", "average from 8:00 to 12:00 at 4000-7700 Angstroms", 
                                   "average from 7:00 and 10:00", "maximum daily"))


knitr::kable(table_data, 
             # col.names = c("linear model formula","R2","adjusted R2"),
             caption = "Table 1: Variables of the dataset airquality.", 
             align =rep("l", 4)) |>
  kable_styling(full_width = FALSE)
```

The threshold for dangerous ozone exposure is according to WHO standards 100 $\mu g/m^3$ which corresponds to 51 ppb. I did not correct for outliers as I am interested in the extreme exposure. I utilized the following libraries and saved the dataset locally to increase knitting speed.

```{r data, message=FALSE, warning=FALSE}
# libraries
library(dplyr); library(lubridate); library(tidyr); library(readr)
library(kableExtra);library(ggplot2);library(slider);library(gridExtra);library(cowplot)
# read downloaded data
airquality <- read_csv("../data/airquality.csv") |>
  
  mutate(
  # add date
  date = as.Date(paste0(as.character(Day),"-", as.character(Month), "-", "1973" ), , format="%d-%m-%Y"),
    
  # add day of week (dow)
  dow  = weekdays(date),
  
  # WHO threshold of 51 ppb
  thresh = if_else(Ozone > 51, "Above", "Below"),
  
  # convert Temperature to celsius 
  Temp = (Temp - 32) * (5/9))

# aggregate data over month and day of the week
airquality_map <- airquality |> 
  group_by(Month, dow) |>
  summarize(
    Ozone = mean(Ozone, na.rm = TRUE), .groups = "drop") |>
  mutate(dow = factor(dow, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                      "Friday", "Saturday", "Sunday")
                      , ordered = TRUE))
```

<br>

I determined the predictability of ozone using the other environmental variables  using linear models.

```{r lm}
# set up data frame
lm_comparison = data.frame(formula = c("Ozone ~ Temp","Ozone ~ Wind","Ozone ~ Solar.R",
                                        "Ozone ~ Temp + Wind","Ozone ~ Temp + Solar.R","Ozone ~ Wind + Solar.R",
                                        "Ozone ~ Temp + Wind + Solar.R"),
                           R2 = rep(NA,7),
                           R2_adj = rep(NA,7))

for (i in 1:nrow(lm_comparison)) {
  
  # linear model
  mod <- lm(as.formula(lm_comparison$formula[i]), data = airquality, na.action = na.exclude)
  
  # save R2 score
  lm_comparison$R2[i] <- round(summary(mod)$r.squared, digits = 3)
  
  # save R2_adj score
  lm_comparison$R2_adj[i] <- round(summary(mod)$adj.r.squared, digits = 3)}
```

<br>

### 3. Results
#### 3.1 Exposure to Ozone

```{r timeseries, echo=FALSE, fig.align='center', fig.width=8, fig.height=5}

ggplot(airquality, aes(x = date, y = Ozone)) +
  geom_abline(slope = 0, intercept = 51, linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point(aes(color = thresh), size = 1.2) +
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  
  
  labs(
    x = "Date",
    y = "Ozone [ppb]",
    title = "Ozone concentration in New York in 1973",
    subtitle = "Excedance of WHO threshold",
    color = "Threshold"
  ) +
  theme_classic()

```

*Figure 1*

Ozone varies greatly from day to day with ozone concentrations ranging from 1 to 168 ppb with a mean of 42 ppb. Higher average concentrations were observed in the summer months (June-August) with on average 55 ppb and lower concentrations in spring (May) with 23 ppb and autumn (September) with 31 ppb (Figure 1). While ozone concentrations were almost exclusively below the WHO threshold of 51 in May and September, they were frequently above this threshold from June to August. However, large fluctuation lead to them being below the threshold from time to time in summer too. 



```{r weeklyplot, echo=FALSE, fig.align='center', fig.width=8, fig.height=4}
ggplot(airquality_map, aes(x = Month, y = dow, fill = Ozone)) +
  geom_tile() +
  scale_fill_gradientn(
    colors = c("skyblue","red", "darkred"),  # Define the color transition
    values = scales::rescale(c(0, 50, 51, max(airquality_map$Ozone, na.rm = TRUE))), 
    na.value = "white"
  ) +
  labs(
    x = "Month",
    y = "Day of the Week",
    fill = "Ozone",
    title = "Ozone Across Months and Weekdays",
    subtitle = "Excedance of WHO threshold"
  ) +
  theme_minimal()

```

*Figure 2*

<br>

Comparing monthly averages, Figure 2 shows that ozone concentrations are highest in July and August with averages above the WHO threshold and lowest in May. Only Friday had consistently across all months decreased ozone concentrations. No other day showed a clear pattern that was visible in all months.

<br>
#### 3.2 Prediction of Ozone
 
The Analysis of the different linear model options of ozone has shown that temperature is the single best predictor of ozone concentrations followed by wind speed and solar radiation. Adding the effect of wind speed and solar radiation increased the linear model accuracy. The best performing linear model included all three predictors. But all of the model specifications are limited in properly modeling ozone as all models have an adjusted R-squared below 0.6.

<br>

```{r table lm, echo=FALSE}
knitr::kable(lm_comparison, 
             col.names = c("Linear model formula","R2","adjusted R2"),
             caption = "Table 1: Fit of the different linear model specifications.", 
             align = c("l","c","c")) |>
  kable_styling(full_width = FALSE)

```
 
The linear association of temperature, the best single predictor, and ozone is displayed in Figure 3. With an adjusted R-squared of 0.483, this model does not properly capture the relationship between ozone and temperature. More fitting could be a non-linear model that better captures the steep rise in ozone concentrations starting at approximately 27 °C.


```{r lm plot, echo = FALSE, fig.align='center', fig.width=8, fig.height=5}

plot1 <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Temperature [°C]",
    y = "Ozone [ppb]",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 16, y = 150, label = "(a)", size = 4, color = "black") 


plot2 <- ggplot(airquality, aes(x = Wind, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Wind Speed [mph]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 5, y = 150, label = "(b)", size = 4, color = "black") 



plot3 <- ggplot(airquality, aes(x = Solar.R, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Solar Radiation [lang]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 90, y = 150, label = "(c)", size = 4, color = "black")




legend_plot <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  # geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    color = "Threshold"
  ) 
           
plot4 <- get_legend(legend_plot)

grid.arrange(plot1, plot2, plot3, plot4, ncol = 4, padding = unit(0, "lines"))
```

*Figure 3*

 
<br>

### 4. Discussion


I hypothesized that ozone concentrations should by higher in summer due to increased solar radiation and I found a clear seasonal pattern with highest concentrations in the summer months. However, I did not find the expected increased ozone concentration on weekdays, possibly due to other processes being involved in ozones production than traffic and industry activity. Against expectations, solar radiation predicted ozone worse than wind speed and temperature, possibly because chemical reactions rely more on temperature and wind speed better accounting for a dilution of polluted air.
In conclusion, New York's population in 1973 was heavily exposed to dangerous ozone concentrations in the summer months by today's standards. 

<br>
