---
title: "5 Air Quality - Report Exercise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,messages=FALSE, warning=FALSE)
rm(list = ls())
```

## Telling a story from data
### Introduction:

Ozone and health impacts, urgency

- Research questions: 
When is human health most threatened?
ppb threshold = 51 ppb (100 ug/m3)
- timeseries plot
- heatmap (months/weekdays)

How well can we predict this threat best based on other variables to extrapolate this threat to other areas?


### Methods
- description of the airquality data set: https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality

method for exposure: WHO threshold

method for precicatbility: lm using Wind/Solar.R/Temp


because we are interested in the outliers we wont correct for them.

### Notes
- At least three statistical metrics from your dataset that aid you in answering your question (e.g., mean values, ranges, etc.).
- At least three publishable figures or tables that show important relationships that aid you in answering your question (e.g., outliers, temporal patterns, scatterplots, etc.).
- Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?Make sure to interpret and discuss your results and hypotheses. Why were you right / why were you wrong?
- max 400 words (aka 700)
- Do I need to scale for lm()? Kinda not but could be nice to try out

```{r data, message=FALSE, warning=FALSE}
# libraries
library(dplyr); library(lubridate); library(tidyr); library(readr); library(stringr); library(purrr); library(caret)

# read downdloaded data
airquality <- read_csv("../data/airquality.csv") |>
  
  mutate(
  # add date
  date = as.Date(paste0(as.character(Day),"-", as.character(Month), "-", "1973" ), , format="%d-%m-%Y"),
    
  # add day of week (dow)
  dow  = weekdays(date),
  
  # WHO threshold of 51 ppb
  thresh = if_else(Ozone > 51, "Above", "Below"),
  
  # convert Temperature to celsius 
  Temp = (Temp - 32) * (5/9))
  
```


TIME SERIE OF OZONE WITH THRESHOLD

```{r timeseries, echo=FALSE}

ggplot(airquality, aes(x = date, y = Ozone)) +
  geom_abline(slope = 0, intercept = 51, linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point(aes(color = thresh), size = 1.2) +
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  
  
  labs(
    x = "Date",
    y = "Ozone [ppb]",
    title = "Ozone concentration in New York in 1973",
    subtitle = "Excedance of WHO threshold",
    color = "Threshold"
  ) +
  theme_classic()

```



MONTHLY WEEKLY CYCLE OF OZONE


```{r weekly}
airquality_map <- airquality |> 
  group_by(Month, dow) |>
  summarize(
    Ozone = mean(Ozone, na.rm = TRUE), .groups = "drop") |>
  mutate(dow = factor(dow, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                      "Friday", "Saturday", "Sunday")
                      , ordered = TRUE))
```


```{r weeklyplot, echo=FALSE}
ggplot(airquality_map, aes(x = Month, y = dow, fill = Ozone)) +
  geom_tile() +
  scale_fill_gradientn(
    colors = c("lightyellow","darkorange", "darkred"),  # Define the color transition
    values = scales::rescale(c(0, 50, 51, max(airquality_map$Ozone, na.rm = TRUE))), 
    na.value = "white"
  ) +
  labs(
    x = "Month",
    y = "Day of the Week",
    fill = "Ozone",
    title = "Ozone Across Months and Weekdays",
    subtitle = "Excedance of WHO threshold"
  ) +
  theme_minimal()

```





 PREDICTABILITY USING DIFFERENT VALUE SPECIFICATIONS
 
 
TABLE for different lm specs
 
```{r lm}

lm_comparison = data.frame(formula = c("Ozone ~ Temp","Ozone ~ Wind","Ozone ~ Solar.R",
                                        "Ozone ~ Temp + Wind","Ozone ~ Temp + Solar.R","Ozone ~ Wind + Solar.R",
                                        "Ozone ~ Temp + Wind + Solar.R"),
                           R2 = rep(NA,7),
                           R2_adj = rep(NA,7))

for (i in 1:nrow(lm_comparison)) {
  
  # linear model
  mod <- lm(as.formula(lm_comparison$formula[i]), data = airquality, na.action = na.exclude)
  
  # save R2 score
  lm_comparison$R2[i] <- round(summary(mod)$r.squared, digits = 3)
  
  # save R2_adj score
  lm_comparison$R2_adj[i] <- round(summary(mod)$adj.r.squared, digits = 3)
}

# sort results by R2_adj
lm_comparison <- lm_comparison |> arrange(desc(R2_adj))

```
 
 
```{r table lm, echo=FALSE}
knitr::kable(lm_comparison, 
             col.names = c("linear model formula","R2","adjusted R2"),
             caption = "Table 1: Fit of the different linear model specifications.", 
             align = c("l","c","c"))

```
 

```{r lm plot, echo = FALSE}


ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") +
  
  labs(
    x = "Temperature [°C]",
    y = "Ozone [ppb]",
    title = "Linear association of daily temperature and ozone concentrations",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +

  annotate("text", x = 16, y = 130, label = "R2 adjusted", size = 5, color = "black") +
  annotate("text", x = 14.6, y = 117, label = lm_comparison$R2_adj[lm_comparison$formula == "Ozone ~ Temp"], size = 4, color = "black")
```

 
 
