---
title: "Sesssion 5 Air Quality - Report Exercise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,messages=FALSE, warning=FALSE)
rm(list = ls())
```

## Telling a story from data

The objective of this report exercise is to practice visualizing data and using statistical metrics to tell a story from data. For this purpose I investigate the [*airquality*](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality) data set with ozone as the target variable which is publicly available and consists of daily measurements of environmental variables from May first to September 30th 1973 in New York (Table 1).

```{r data, message=FALSE, warning=FALSE}
# load libraries
library(dplyr); library(lubridate); library(tidyr); library(readr);library(cowplot)
library(kableExtra);library(ggplot2);library(slider);library(gridExtra)

# read data (saved locally)
airquality <- read_csv("../data/airquality.csv") |>
  
  mutate(
  # add date
  date = as.Date(paste0(as.character(Day),"-", as.character(Month), "-", "1973" ), , format="%d-%m-%Y"),
    
  # add day of week (dow)
  dow  = weekdays(date),
  
  # convert Temperature to degree Celsius 
  Temp = (Temp - 32) * (5/9)
  )
```

```{r table data, echo= FALSE}
table_data = data.frame(Variable = c("Ozone", "Solar radiation", "Wind speed", "Temperature"),
                        Unit = c("ppb", "langleys", "mph", "°C"),
                        Site = c("Roosevelt Island", "Central Park", "LaGuardia Airport", "LaGuardia Airport"),
                        Method = c("average from 13:00 to 15:00", "average from 8:00 to 12:00 at 4000-7700 Angstroms", 
                                   "average from 7:00 and 10:00", "maximum daily"))


knitr::kable(table_data, 
             caption = "Table 1: Variables of the dataset airquality.", 
             align =rep("l", 4)) |>
  kable_styling(full_width = FALSE, position = "center") |>
  scroll_box(width = "100%")
```

<br>

### 1. Introduction
Ground level ozone is a threat to human health and leads to higher mortality and hospitalization rates $^1$.  Ozone can form when pollutants from traffic or industry like nitrous oxide react with solar radiation $^{2,3}$. Therefore, ozone is highly abundant in populated and polluted urban areas like New York. The degree of the exposure to ozone and their health consequences depend on the ozone concentration and the timely distribution and variation of ozone. In order to decrease the potential health effect of dangerous ozone exposure, ozone must be measured or sufficiently predicted from other more common meteorological variables. In this report exercise, I will investigate the seasonal and weekly patterns of ozone concentrations and determine possible predictors like wind speed, solar shortwave radiation and temperature of dangerous ozone exposure. I hypothesize that ozone concentrations are higher in the summer due increased solar radiation and during the week due to increased traffic and industry activity. Furthermore, ozone should be best predicted by solar radiation as it is integral to its formation. 

<br>

### 2. Methods
When investigating the health effects of ozone, a health impacting concentration threshold must be defined. The threshold for dangerous ozone exposure is according to WHO standards 100 $\mu g/m^3$ which corresponds to 51 ppb $^4$. Because I am interested in the exposure to ozone which is most severe in extremes I did not correct for outliers.

```{r data thrsh, message=FALSE, warning=FALSE}
airquality <-  airquality |>
  
  # WHO threshold of 51 ppb
  mutate(thresh = if_else(Ozone > 51, "Above", "Below")
         )
```

In order to investigate the weekly and monthly cycle of ozone, I aggregated the data by the day of the week and month.
```{r data aggregate, message=FALSE, warning=FALSE}
airquality_dow <- airquality |>
  
  group_by(Month, dow) |>
  
  summarize(
    ozone = mean(Ozone, na.rm = TRUE), .groups = "drop"
    ) |>
  
  mutate(
    dow = factor(dow, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                      "Friday", "Saturday", "Sunday"),
                 ordered = TRUE),
    thresh = ozone > 51
    )
```

The predictability of ozone using the other environmental variables was determined using linear models with single predictors.

```{r lm}
# empty data frame
lm_comparison = data.frame(formula = c("Ozone ~ Temp",
                                       "Ozone ~ Wind",
                                       "Ozone ~ Solar.R"),
                           R2 = rep(NA,3),
                           R2_adj = rep(NA,3)
                           )

# loop through model specifications
for (i in 1:nrow(lm_comparison)) {
  
  # linear model
  mod <- lm(as.formula(lm_comparison$formula[i]), data = airquality, na.action = na.exclude)
  
  # save R2 score
  lm_comparison$R2[i] <- round(summary(mod)$r.squared, digits = 3)
  
  # save R2_adj score
  lm_comparison$R2_adj[i] <- round(summary(mod)$adj.r.squared, digits = 3)
  }
```

<br>

### 3. Results
#### 3.1 Seasonality of ozone

```{r timeseries, echo=FALSE, fig.align='center', fig.width=7, fig.height=3.5}

ggplot(airquality, aes(x = date)) +
  
  geom_abline(slope = 0, intercept = 51, linetype = "dashed") +
  
  geom_point(aes(y=Ozone),linewidth = 1) +
  geom_line(aes(y=Ozone),linewidth = 1) +
  
  labs(
    x = "Date",
    y = "Ozone [ppb]",
    title = "",
    subtitle = "",
    color = "threshold"
  ) +
  theme_classic()

```
<center>
*Figure 1: Ozone concentration in New York in 1973 relative to the WHO threshold of 551 ppb (dashed line).*
</center>

<br>

Ozone varies greatly from day to day with ozone concentrations ranging from 1 in May to 168 in August ppb with an average daily concentration of 42 ppb and a median concentration of 32 ppb. Higher average concentrations were observed in the summer months (June-August) with on average 55 ppb and lower concentrations in spring (May) with 23 ppb and autumn (September) with 31 ppb (Figure 1). While ozone concentrations were almost exclusively below the WHO threshold of 51 in May and September, they were frequently above this threshold from June to August. However, large fluctuation lead to them being below the threshold from time to time in summer too. 

<br>

```{r weeklyplot, echo=FALSE, fig.align='center', fig.width=7, fig.height=3.5}
ggplot(airquality_dow, aes(x = Month, y = dow, fill = ozone)) +
  geom_tile() +
    scale_fill_gradientn(
    colors = c("gold", "red3"),  # Define the color transition
    values = scales::rescale(c(0, 85)),
    na.value = "white"
  ) +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Day of the Week",
    fill = "Ozone",
    title = "",
    subtitle = ""
  ) +
  geom_tile(data = filter(airquality_dow, ozone > 51),
          color = "black", fill = NA, size = 0.65,
          linetype = "dashed")




```
<center>
*Figure 2: Ozone Across Months and Weekdays with respect to the WHO treshold of 51 ppb*
</center>

<br>

Comparing monthly averages, Figure 2 shows that ozone concentrations were highest in July and August with multiple daily averages above the WHO threshold. No other other month showed on any weekday an daily average above the threshold of 51 ppb. Only Friday had consistently across all months decreased ozone concentrations. No other day showed a clear pattern that was visible in all months.

<br>
#### 3.2 Prediction of Ozone
 
The Analysis of the different linear model options of ozone has shown that temperature is the single best predictor of ozone concentrations followed by wind speed and solar radiation. Adding the effect of wind speed and solar radiation increased the linear model accuracy. The best performing linear model included all three predictors. But all of the model specifications are limited in properly modeling ozone as all models have an adjusted R-squared below 0.6.

<br>

```{r table lm, echo=FALSE}
knitr::kable(lm_comparison, 
             col.names = c("Linear model formula","R2","adjusted R2"),
             caption = "Table 1: Fit of the different linear model specifications.", 
             align = c("l","c","c")) |>
  kable_styling(full_width = FALSE)

```
 
The linear association of temperature, the best single predictor, and ozone is displayed in Figure 3. With an adjusted R-squared of 0.483, this model does not properly capture the relationship between ozone and temperature. More fitting could be a non-linear model that better captures the steep rise in ozone concentrations starting at approximately 27 °C.


```{r lm plot, echo = FALSE, fig.align='center', fig.width=8, fig.height=5}

plot1 <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Temperature [°C]",
    y = "Ozone [ppb]",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 16, y = 150, label = "(a)", size = 4, color = "black") 


plot2 <- ggplot(airquality, aes(x = Wind, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Wind Speed [mph]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 5, y = 150, label = "(b)", size = 4, color = "black") 



plot3 <- ggplot(airquality, aes(x = Solar.R, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    x = "Solar Radiation [lang]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +

  annotate("text", x = 90, y = 150, label = "(c)", size = 4, color = "black")




legend_plot <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red", "Below" = "skyblue3"), drop = TRUE, na.translate = FALSE) +
  # geom_smooth(formula = y ~ x + 0, method = "lm", se = FALSE, color = "black") + 
  
  labs(
    color = "Threshold"
  ) 
           
plot4 <- get_legend(legend_plot)

grid.arrange(plot1, plot2, plot3, plot4, ncol = 4, padding = unit(0, "lines"))
```

*Figure 3*

 
<br>

### 4. Discussion


I hypothesized that ozone concentrations should by higher in summer due to increased solar radiation and I found a clear seasonal pattern with highest concentrations in the summer months. However, I did not find the expected increased ozone concentration on weekdays, possibly due to other processes being involved in ozones production than traffic and industry activity. Against expectations, solar radiation predicted ozone worse than wind speed and temperature, possibly because chemical reactions rely more on temperature and wind speed better accounting for a dilution of polluted air.
In conclusion, New York's population in 1973 was heavily exposed to dangerous ozone concentrations in the summer months by today's standards. 

<br>

### References

[1] Hubbell BJ, Hallberg A, McCubbin DR, Post E. Health-related benefits of attaining the 8-hr ozone standard. Environ Health Perspect. 2005 Jan;113(1):73-82. doi: 10.1289/ehp.7186. PMID: 15626651; PMCID: PMC1253713.

[2] Shan, Danying & Du, Zhenyu & Zhang, Ting & Zhang, Xiulan & Cao, Guan & Liu, Zirui & Yao, Zhixiao & Tang, Ka & Liang, Siyuan. (2023). Variations, sources, and effects on ozone formation of VOCs during ozone episodes in 13 cities in China. Frontiers in Environmental Science. 10. 1084592. 10.3389/fenvs.2022.1084592.

[3] Hu, C., Kang, P., Jaffe, D. A., Li, C., Zhang, X., Wu, K., & Zhou, M. (2021). Understanding the impact of meteorology on ozone in 334 cities of China. In Atmospheric Environment (Vol. 248, p. 118221). Elsevier BV. doi.org/10.1016/j.atmosenv.2021.118221

[4] WHO global air quality guidelines. Particulate matter (PM2.5 and PM10), ozone,
nitrogen dioxide, sulfur dioxide and carbon monoxide. Geneva: World Health Organization; 2021. Licence:
CC BY-NC-SA3.0 IGO.


<br>
