---
title: "Sesssion 5 Air Quality - Report Exercise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,messages=FALSE, warning=FALSE)
rm(list = ls())
```

## Telling a story from data

The objective of this report exercise is to practice visualizing data and using statistical metrics to tell a story from data. For this purpose I investigated the [*airquality*](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality) data set with ozone as the target variable, which is publicly available and consists of daily measurements of environmental variables from May first to September 30th 1973 in New York (Table 1).

```{r data, message=FALSE, warning=FALSE}
# load libraries
library(lubridate);library(cowplot);library(kableExtra);library(slider);library(gridExtra);library(tidyverse)

# read data
airquality <- read_csv("../data/airquality.csv") |>
  
  mutate(
  # add date
  date = as.Date(paste0(as.character(Day),"-", as.character(Month), "-", "1973" ), , format="%d-%m-%Y"),
    
  # add day of week (dow)
  dow  = weekdays(date),
  
  # convert Temperature to degree Celsius 
  Temp = (Temp - 32) * (5/9)
  )
```

```{r table data, echo= FALSE}
table_data = data.frame(Variable = c("Ozone", "Solar radiation", "Wind speed", "Temperature"),
                        Unit = c("ppb", "langleys", "mph", "°C"),
                        Site = c("Roosevelt Island", "Central Park", "LaGuardia Airport", "LaGuardia Airport"),
                        Method = c("average from 13:00 to 15:00", "average from 8:00 to 12:00 at 4000-7700 Angstroms", 
                                   "average from 7:00 and 10:00", "maximum daily"))


knitr::kable(table_data, 
             caption = "Table 1: Variables of the airquality dataset.", 
             align =rep("l", 4)) |>
  kable_styling(full_width = FALSE, position = "center") |>
  scroll_box(width = "100%")
```

<br>

### 1. Introduction
Ground level ozone is a threat to human health and leads to higher mortality and hospitalization rates $^1$.  Ozone can form when pollutants from traffic or industry, like nitrous oxide, react with solar radiation $^{2,3}$. Therefore, ozone is highly abundant in populated and polluted urban areas like New York. The degree of the exposure to ozone and their health consequences depend on the ozone concentration, its timely distribution, and the variation of ozone. In order to decrease the potential health effects of dangerous ozone exposure, ozone must be monitored or sufficiently predicted from other more common meteorological variables. In this report exercise, I will investigate the seasonal and weekly patterns of ozone concentrations and determine possible single predictors of ozone concentration like wind speed, solar shortwave radiation, and temperature. I hypothesize that ozone concentrations are higher in the summer due to increased solar radiation and from Monday to Friday due to increased traffic and industry activity. Furthermore, ozone should be best predicted by solar radiation, as it is integral to its formation. 

<br>

### 2. Methods
When investigating the health effects of ozone, a health impacting concentration threshold must be defined. The threshold for dangerous ozone exposure is according to WHO standards 100 $\mu g/m^3$ over 8 hours, which corresponds to 51 ppb $^4$. The data was not investigated for outliers in its distribution, due to the extremes being of special interest in an exposure study.

```{r data thrsh, message=FALSE, warning=FALSE}
airquality <-  airquality |>
  
  # WHO threshold of 51 ppb
  mutate(
    thresh = if_else(Ozone > 51, "Above", "Below")
         )
```

In order to investigate the weekly and monthly cycle of ozone, I aggregated the data by the day of the week and month.
```{r data aggregate, message=FALSE, warning=FALSE}
airquality_dow <- airquality |>
  
  group_by(Month, dow) |>
  
  summarize(
    ozone = mean(Ozone, na.rm = TRUE), .groups = "drop"
    ) |>
  
  mutate(
    dow = factor(dow, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                      "Friday", "Saturday", "Sunday"),
                 ordered = TRUE),
    thresh = ozone > 51
    )
```

The predictability of ozone using the other environmental variables was determined using linear models with single predictors.

```{r lm}
# empty data frame
lm_comparison = data.frame(formula = c("Ozone ~ Temp",
                                       "Ozone ~ Wind",
                                       "Ozone ~ Solar.R"),
                           R2 = rep(NA,3)
                           )

# loop through model specifications
for (i in 1:nrow(lm_comparison)) {
  
  # linear model
  mod <- lm(as.formula(lm_comparison$formula[i]), data = airquality, na.action = na.exclude)
  
  # save R2 score
  lm_comparison$R2[i] <- round(summary(mod)$r.squared, digits = 3)
  }
```

<br>

### 3. Results
#### 3.1 Seasonality of ozone

```{r timeseries, echo=FALSE, fig.align='center', fig.width=6.5, fig.height=3.5}

ggplot(airquality, aes(x = date)) +
  
  geom_abline(slope = 0, intercept = 51, linetype = "dashed") +
  
  geom_point(aes(y=Ozone),linewidth = 1) +
  geom_line(aes(y=Ozone),linewidth = 1) +
  
  labs(
    x = "Date",
    y = "Ozone [ppb]",
    title = "",
    subtitle = "",
    color = "threshold"
  ) +
  theme_classic()

```
<center>
*Figure 1: Daily ozone concentration in New York in 1973 relative to the WHO threshold of 51 ppb (dashed line).*
</center>

<br>

Ozone varied greatly from day to day, with ozone concentrations ranging from 1 ppb in May to 168 ppb in August with an average daily concentration of 42 ppb, a median concentration of 32 ppb, and a standard deviation of 33 ppb. Higher average concentrations were observed in the summer months (June-August) with on average 55 ppb which is above the WHO threshold, and lower concentrations in spring (May) with 23 ppb and autumn (September) with 31 ppb (Figure 1). While ozone concentrations were almost exclusively below the WHO threshold of 51 in May and September, they were frequently severely above this threshold from June to August.

<br>

```{r weeklyplot, echo=FALSE, fig.align='center', fig.width=7, fig.height=3.5}
ggplot(airquality_dow, aes(x = Month, y = dow, fill = ozone)) +
  geom_tile() +
    scale_fill_gradientn(
    colors = c("gold", "red3"),  # Define the color transition
    values = scales::rescale(c(0, 85)),
    na.value = "white"
  ) +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Day of the week",
    fill = "Ozone",
    title = "",
    subtitle = ""
  ) +
  geom_tile(data = filter(airquality_dow, ozone > 51),
          color = "black", fill = NA, size = 0.65,
          linetype = "dashed")




```
<center>
*Figure 2: Ozone aggregated across months and weekdays with days exceeding 51 ppb indicated by dashed borders.*
</center>

<br>
Figure 2 shows that all days of the week exceeding 51 ppb on average were in July and August, with 5 and 6 days, respectively. In July, the highest average concentrations were measured on Wednesday and in August on Saturday. July and August shared no pattern in exceeding 51 ppb of ozone. Taking all other months into account, weekdays and weekends were not consistent in their differences, with May showing higher concentrations during weekdays and June showing higher concentrations over the weekend. Fridays had lower ozone concentrations compared to all other weekdays across all months.

<br>

#### 3.2 Prediction of Ozone
 
From the three single predictors, temperatures linear model performed best at predicting ozone with an $R^2$ of 0.488 which indicates that roughly 50% of the variation of ozone can be modeled with just temperature. The linear models of wind speed and solar radiation performed worse and were only poorly able to predict ozone concentrations (Table 2).

<br>

```{r table lm, echo=FALSE}
knitr::kable(lm_comparison, 
             col.names = c("Linear model formula","R2"),
             caption = "Table 2: Fit of the different linear model specifications.", 
             align = c("l","c")) |>
  kable_styling(full_width = FALSE)

```
 

Comparing the linear models visually, the temperature-ozone association indicates a non-linear relationship and shows increasing ozone concentrations with increasing temperature. There is a steep incline in ozone concentration at high temperatures where the model underestimates ozone concentrations (Figure 3). Wind speed as a predictor had an inverse association with ozone, as the highest concentrations occurred in low wind speed conditions. Wind speeds model also had wider confidence intervals than the temperature model, which again indicates poorer predictability. There is no clear visual pattern between ozone and solar radiation, supporting the very poor predictability of ozone shown in Table 2.

```{r lm plot, echo = FALSE, fig.align='center', fig.width=8, fig.height=5}

plot1 <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh), size = 1.2 )+
  scale_color_manual(values = c("Above" = "red2", "Below" = "gold"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x, method = "lm", se = TRUE, color = "black", linetype = 1, size = 0.8) + 
  
  labs(
    x = "Temperature [°C]",
    y = "Ozone [ppb]",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  ylim(c(0,150)) +

  annotate("text", x = 16, y = 150, label = "A", size = 4, color = "black") 


plot2 <- ggplot(airquality, aes(x = Wind, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red2", "Below" = "gold"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x, method = "lm", se = T, color = "black", linetype = 1, size = 0.8) + 
  
  labs(
    x = "Wind speed [mph]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  ylim(c(0,150))+
  annotate("text", x = 4, y = 150, label = "B", size = 4, color = "black") 



plot3 <- ggplot(airquality, aes(x = Solar.R, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red2", "Below" = "gold"), drop = TRUE, na.translate = FALSE) +
  geom_smooth(formula = y ~ x, method = "lm", se = T, color = "black", linetype = 1, size = 0.8) + 
  
  labs(
    x = "Solar radiation [lang]",
    y = "",
    title = "",
    subtitle = "",
    color = "Threshold"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  ylim(c(0,150))+
  annotate("text", x = 40, y = 150, label = "C", size = 4, color = "black")




legend_plot <- ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(aes(color = thresh ), size = 1.2)+
  scale_color_manual(values = c("Above" = "red2", "Below" = "gold"), drop = TRUE, na.translate = FALSE) +
  
  labs(
    color = "Threshold"
  ) 
           
plot4 <- get_legend(legend_plot)

grid.arrange(plot1, plot2, plot3, plot4, ncol = 4, padding = unit(0, "lines"))
```
<center>
*Figure 3: Scatterplots of temperature, wind speed, and solar radiation as predictors of ozone and their linear models as dashed lines.*
</center>
 
<br>

### 4. Discussion


I hypothesized that ozone concentrations should be higher in the summer due to increased solar radiation, and I found a clear seasonal pattern with the highest concentrations in the summer months. However, I did not find the expected increased ozone concentration on weekdays, possibly due to other processes being involved in ozone production and decay than traffic and industry activity. Against expectations, solar radiation predicted ozone worse than wind speed and temperature, possibly because chemical reactions rely more on temperature and wind speed better accounting for a dilution of polluted air. Temperature performed best at predicting ozone concentrations but was not sufficiently accurate to serve as a sole predictor in health monitoring programs. In conclusion, New York's population in 1973 was heavily exposed to dangerous ozone concentrations in the summer months, which often coincided with high temperatures. 

<br>

### References

[1] Hubbell BJ, Hallberg A, McCubbin DR, Post E. Health-related benefits of attaining the 8-hr ozone standard. Environ Health Perspect. 2005 Jan;113(1):73-82. doi: 10.1289/ehp.7186. PMID: 15626651; PMCID: PMC1253713.

[2] Shan, Danying & Du, Zhenyu & Zhang, Ting & Zhang, Xiulan & Cao, Guan & Liu, Zirui & Yao, Zhixiao & Tang, Ka & Liang, Siyuan. (2023). Variations, sources, and effects on ozone formation of VOCs during ozone episodes in 13 cities in China. Frontiers in Environmental Science. 10. 1084592. 10.3389/fenvs.2022.1084592.

[3] Hu, C., Kang, P., Jaffe, D. A., Li, C., Zhang, X., Wu, K., & Zhou, M. (2021). Understanding the impact of meteorology on ozone in 334 cities of China. In Atmospheric Environment (Vol. 248, p. 118221). Elsevier BV. doi.org/10.1016/j.atmosenv.2021.118221

[4] WHO global air quality guidelines. Particulate matter (PM2.5 and PM10), ozone,
nitrogen dioxide, sulfur dioxide and carbon monoxide. Geneva: World Health Organization; 2021. Licence:
CC BY-NC-SA3.0 IGO.


<br>
