---
title: "Session 4 Data wrangling - Report Exerise"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

## Analyzing changes in soil organic matter during elevated CO2 experiments
 
In this report exercise, I will practice Open Science principles by converting a badly set up published data set into a data set that is tidy and consequently allows for a reproducible workflow. Afterward, I will investigate the effect of increased atmospheric CO2 concentrations on soil organic matter using the previously cleaned data set.

The data is taken from *Groenigen, Kees Jan van, Xuan Qi, Craig W. Osenberg, Yiqi Luo, and Bruce A. Hungate. “Faster Decomposition Under Increased Atmospheric CO2 Limits Soil Carbon Storage.” Science 344, no. 6183 (May 2, 2014): 508–9. [https://doi.org/10.1126/science.1249534](https://doi.org/10.1126/science.1249534).* It is saved in the /dataraw folder. Before my investigation I cleaned it manually using the following steps and saved it in the /data folder.

* deleting empty rows,
* deleting the first 3 rows that were empty,
* replacing spaces with "_" in column names, 
* filling the column *Experiments* with corresponding information,
* converting the column depth to a machine readable numeric value, and
* converting the entries of the column *Sample dates* to a homogeneous date format.

<br>

```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(dplyr); library(lubridate); library(tidyr); library(readr);library(kableExtra);library(ggplot2)

# load clean data
data <- read_csv("../data/groenigen_DBS1.csv", locale = locale(encoding = "UTF-8"))[,1:11]|>
  mutate(Sample_date = dmy(Sample_date))  # convert date column to date class
```

<br>

```{r data table, echo=FALSE}
knitr::kable(data[1:5,1:9], 
             col.names = c("Experiment","Citation","Depth","Date","Time","AmbientCO2","IncreasedCO2", "nAmbient", "nIncreased"),
             caption = "Table 1: Data overview without the columns \"Descriptions\" and \"Value treatment\".", 
             align = c("l","l","c","l",rep("c",5))) |>
  kable_styling(full_width = FALSE, position = "center") |>
  scroll_box(width = "100%")
```

### 1. Introduction
In the era of climate change, it is vital to understand the effects of increased atmospheric CO2 concentrations have on ecosystems. Soil stores carbon in the form of organic and inorganic matter. Organic soil matter is cut-up dead biomass primarily contributed by plants. The growth of plants is regulated by their photosynthesis rate which can be limited or enhanced by the abundance of atmospheric CO2. 

In this report I will investigate whether increased atmospheric CO2 concentrations lead to an increasing storage of carbon in soil. I hypothesize that (i) higher CO2 concentrations lead to a short term increase in soil carbon storage and (ii) that soil carbon will eventually reach an equilibrium at those higher concentrations and not rise further.  

<br>

### 2. Methods
I analyzed the effect of atmospheric CO2 concentrations on soil carbon concentrations by summarizing multiple studies which compared soil samples which were divided into samples from an environment with ambient CO2 concentrations and with increased CO2 concentrations.

The data was provided by *Groenigen et al. (2014)* who collected existing literature on the responses of elevated CO2 concentrations on soil organic matter. I identified whether there is an effect of this treatment using the log-response ratio comparing ambient CO2 and elevated CO2 soil organic matter across all experiments. However, I distinguished between different time phases after the start of the experiment as these could indicate a timely development of the response to the elevated CO2 concentrations. The phases are:

* early: start of the experiment and up to 3 years
* mid: between 3 and 6 years after the start of the experiment
* late: more than 6 years since the start of the experiment

The formula used for calculating the log-response ratio RR is the following with $x$ indicating the concentration of soil organic carbon in the two scenarios:
$$RR = ln( \frac{x_{elevated}}{x_{ambient}} )$$


I aggregated the log-response ratios across all experiments using the following steps:

1. Calculate the averages of ambient and increased CO2 soil carbon concentrations for every experiment and every phase,
2. calculate the log-response ratios of all existing experiment-phase pairs, and
3. then average over all values of corresponding to the same phase.
 
 
The averaging of the log-response ratios had to be adjusted, because not all experiments had measured over enough time to represent all of our phases or were missing data. Consequently, we had to exclude all response ratios from the averaging process that were either not a number (NaN) and exactly 0 because one or both of the division partners were missing. A response ratio of 0 could also mean that there is exactly no effect of the increased CO2 concentrations. However, having precisely no difference at all is highly unlikely and could also result out of measurement errors.
<br>

```{r RR}
#  aggregate across experiments and phases
data_aggregated <- data |> 
  
  # select columns that we investigate
  select(Experiment, Time_years, mean_ambient_CO2, mean_increased_CO2) |>
  
  # group by experiment
  group_by(Experiment) |>
  
  # 1. summarize across ambient and increased CO2 values in 3 phases
  summarize(
    mean_ambient_CO2_early = mean(mean_ambient_CO2[Time_years < 3], na.rm = T),                        # early
    mean_ambient_CO2_mid   = mean(mean_ambient_CO2[Time_years >= 3 & Time_years <= 6], na.rm = T),     # mid
    mean_ambient_CO2_late  = mean(mean_ambient_CO2[Time_years > 6], na.rm = T),                        # late
    
    mean_increased_CO2_early = mean(mean_increased_CO2[Time_years < 3], na.rm = T),                    # early
    mean_increased_CO2_mid   = mean(mean_increased_CO2[Time_years >= 3 & Time_years <= 6], na.rm = T), # mid
    mean_increased_CO2_late  = mean(mean_increased_CO2[Time_years > 6], na.rm = T),                    # late
    
    # 2. calculate RR for every parallel observation 
    RR_early = log(mean_increased_CO2_early / mean_ambient_CO2_early),   # early
    RR_mid   = log(mean_increased_CO2_mid / mean_ambient_CO2_mid),       # mid
    RR_late  = log(mean_increased_CO2_late / mean_ambient_CO2_late)      # late
    )

# 3. average the response ratios for every phase
data_RR_means <- data_aggregated |>
  
  summarize(
    ean_RR_early = mean(RR_early[!is.nan(RR_early) & RR_early != 0], na.rm = T),         # early
    mean_RR_mid   = mean(RR_mid[!is.nan(RR_mid) & RR_early != 0], na.rm = T),            # mid
    mean_RR_late  = mean(RR_late[!is.nan(RR_late) & RR_early != 0], na.rm = T)           # late
  )
```


<br>

### 3. Results

There are clear absolute differences between experiments and their recorded average soil carbon concentrations (Table 2). From the 5 experiments shown in Table 2, there is also no clear pattern visible between early, mid and late phase soil carbon concentrations. Response ratios varied from phase to phase but also experiment to experiment. Multiple experiments did not measure soil organic matter for all of our defined phases. Consequently, 32% of all response ratios are missing. 

```{r data aggregated table, echo=FALSE}
# find out the percentage of missing RR of all phases
# nan_percentage <- sum(is.nan(as.matrix(data_aggregated[,8:10]))) / prod(dim(data_aggregated[,8:10])) * 100
# -> 32%

# round values and replace NaNs with "-"
data_aggregated_table <- data_aggregated |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  mutate(across(everything(), ~ifelse(is.nan(.), "-", .)))

knitr::kable(data_aggregated_table[1:5,],  
             col.names = c("Experiment","ACO2_early","ACO2_mid","ACO2_late","ICO2_early","ICO2_mid","ICO2_late", "RR_early", "RR_mid", "RR_late"), 
             digits = 2, 
             caption = "Table 2: Cutout of the aggregated data comparing ambient CO2 concentrations (ACO2) and increased CO2 concentrations (ICO2) and their response ratios (RR) in three phases.") |>
  kable_styling(full_width = FALSE, position = "center") |>
  scroll_box(width = "100%")

```


<br>

Through averaging over all experiments we get a more representative picture of the effect of increased atmospheric carbon on soil organic matter (Table 3). The aggregated response ratios show higher soil carbon concentrations in the increased atmospheric CO2 scenario in the first 3 years. In the second phase (3-6 years), we have a negative response ratio indicating less carbon in the elevated CO2 scenario compared with soil from ambient conditions. After 6 years, the difference between elevated CO2 soil and ambient CO2 soil decreases again but stays slightly negative, indicating slightly higher soil carbon concentrations in the ambient scenario.

```{r RR table, echo=FALSE}
knitr::kable(data_RR_means,  
             col.names = c("RR early phase","RR mid phase","RR late phase"), digits = 4,
             caption = "Table 3: Response ratios to comparing ambient to increased CO2 concentrations.")  |>
  kable_styling(full_width = FALSE, position = "center") |>
  scroll_box(width = "100%")

```






<br>

### 4. Discussion
I hypothesized that elevated CO2 concentrations could lead to a short term increase in soil carbon storage, which would reach an equilibrium with higher concentrations in the elevated CO2 scenario. I did find soil carbon to increase at first due to the increased abundance of atmospheric CO2, however, after some time the soil carbon decreased again below levels of the ambient CO2 scenario. Interestingly, the soil stores less carbon with higher atmospheric carbon availability after some time passing. The authors of *Groenigen et al. (2014)* argue that this could be due to an increase in the microbial activity that decomposes soil organic carbon, which I did not take into account while formulating my hypotheses. These microbes respirate soil organic carbon faster than it is produced in elevated CO2 conditions, leading to an equilibrium below the levels of the ambient scenario after a first spike of soil carbon concentrations. The global increase in temperature could even further exacerbate this shift to a decrease of soil organic carbon storage. This could have dire consequences as soils would not act as a carbon sink that is able to keep up with rising CO2 concentrations, according to my results.

<br>
