---
title: "2 Getting started - Workspace"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

### 1. Dimensions of a circle
I chose to write a *function* to get the dimensions of a circle based on the radius as a text output. This enables me to try out different radii efficiently or just try out random ones. 

```{r dimensions}
# function
circle <- function(r){
  
  # formula for area of the circle
  area = pi * r^2
  
  # formula for circumference of the circle
  circ = 2 * pi * r

  # text output for both
  print(paste0("the area of the circle with the radius of ", r, " is: ",
               round(area, digits = 2)))

  print(paste0("the circumference of the circle with the radius of ", r, " is: ",
               round(circ, digits=2)))
}

# call of the function with 
circle(round(runif(1,1,10), digits = 2))
```
<br>

### 2. Sequence of numbers
The most efficient way to create a sequence of 5 numbers from 0 to pi using 5 number is the *seq()* function.

```{r sequence}
vec <- seq(0, pi ,length = 5)
print(vec)
```
<br>

### 3. Gauss sum
Sadly I was not as intelligent as young Carl in primary school. But luckily I don't have to. I can use the *sum()* function. I doubt that Carl was as fast as my 3.2 GHz processor.
```{r gauss}
print(paste0("The answer is  ",sum(1:100), " Mr. Gauss"))
```
<br>

### 4. Magic trick algorithm
Somebody showed this magic trick to me in middle school, truly fascinating to a 12 year old! Again a function ensures a fast cycle of dopamine rushes.

```{r magic}
magic <- function(x){
  
  #save original value
  b <- x
  
  print(paste0("Your chose the number: ", b, ". Abrakadabra!!!"))
  print("..plus 1..times 2..plus 4..divided by 2..minus the original number..")
  
  # sequence of maths
  x <- x + 1
  x <- x * 2
  x <- x + 4
  x <- x / 2
  x <- x - b

  print(paste0("Your new number is: ", x, ". Try one more time to see the magic!"))
}

# call magic
magic(666)
```
<br>

### 5. Vectors
R can be used to summarize and and get a first look at river data.

```{r vectors}
# data
data_rivers <- datasets::rivers

# class of the data
class(data_rivers)

# length of the data
length(data_rivers)

# mean, minimum and maximum
mean(data_rivers); min(data_rivers);max(data_rivers) 

# median and 33th percentile
quantile(data_rivers, c(.5,.33))
```
<br>

### 6. Data frames
Lets find the geographic location of the largest earthquake of the *quakes* dataset!

```{r quakes}
# data
data_quakes <- datasets::quakes

# dimensions
dim(data_quakes)

# largest magnitude
magnitude_max <- max(data_quakes$mag)

# location
paste0("location: latitude: ", data_quakes$lat[data_quakes$mag == magnitude_max]," and longitude: " , data_quakes$lon[data_quakes$mag == magnitude_max])
```

<br>









